---
title: "Aim 1: phase DE"
output: html_document
date: "2025-03-03"
---

```{r}
library(tidyverse)
library(lattice)
library(clusterProfiler)
library(enrichplot)
library(edgeR)
```

#load count data
```{r}
load("anubis_featurecounts.Rdata")
counts <- as.data.frame(anubisfc$counts)
colnames(counts)<-gsub(".anubis.useqs.merged.bam", "", colnames(counts)) 
```

#Find best phase samples
```{r}
CountsperSampleType<-read.csv("DE_sample_metadata.csv")

NotPostcop<-subset(CountsperSampleType, Postcop == "N")

PreFertile<-subset(NotPostcop, phasecode == "PreF") 
PreFertile_ordered <- PreFertile[order(PreFertile$uniquely_mapped_total, decreasing=TRUE), ]
PreFertile_top8<-PreFertile_ordered[1:8,]

Fertile<-subset(NotPostcop, phasecode == "F")
Fertile_ordered <- Fertile[order(Fertile$uniquely_mapped_total, decreasing=TRUE), ]
Fertile_top8<-Fertile_ordered[1:8,]

PostFertile<-subset(NotPostcop, phasecode == "PostF")
PostFertile_ordered <- PostFertile[order(PostFertile$uniquely_mapped_total, decreasing=TRUE), ]
PostFertile_top8<-PostFertile_ordered[1:8,]

NonFertile<-subset(NotPostcop, phasecode == "NonF")
NonFertile_ordered <- NonFertile[order(NonFertile$uniquely_mapped_total, decreasing=TRUE), ]
NonFertile_top8<-NonFertile_ordered[1:8,]

Top32<-do.call(rbind, list(PreFertile_top8, Fertile_top8, PostFertile_top8, NonFertile_top8))

VagNoPostcoptop32 <- CountsperSampleType %>%
  filter(Sample_ID %in% Top32$Sample_ID)

VagNoPostcoptop32Phase <- as.vector(VagNoPostcoptop32$phasecode)
```

#filter counts for phase samples
```{r}
countsnopostcop_top32 <- counts[,colnames(counts) %in% Top32$Sample_ID]
```

#Create DGEList
```{r}
DGEList= NULL
DGEList <- DGEList(counts=countsnopostcop_top32, genes=anubisfc$annotation[,c("GeneID","Length")], group=factor(VagNoPostcoptop32Phase))
```

#Remove low expressed genes (>10cpm in at least half of samples)
```{r}
keep <- rowSums(cpm(DGEList) > 10) >= 16
FilteredDGEList <- DGEList[keep,]
```

#Remove genes starting with "LOC"
```{r}
FilteredDGEList$counts<- FilteredDGEList$counts[!grepl("LOC", rownames(FilteredDGEList$counts)), ]
FilteredDGEList$genes<- FilteredDGEList$genes[!grepl("LOC", rownames(FilteredDGEList$genes)), ]
```

#Reset the library sizes
```{r}
FilteredDGEList$samples$lib.size <- colSums(FilteredDGEList$counts)
```


################
# Differential expression analyses
################


#Create vectors with metadata information
```{r}
VagNoPostcoptop32$phasecode <- relevel(as.factor(VagNoPostcoptop32$phasecode), ref = "NonF")
VagNoPostcoptop32$male_status<-gsub(" ", "", VagNoPostcoptop32$male_status)
MalePresence <- as.vector(VagNoPostcoptop32$male_status)
RIN <- as.vector(VagNoPostcoptop32$RIN_RQN.y)
FemaleID <- as.vector(VagNoPostcoptop32$FemaleID)
```

#Run model
```{r}
DGE_counts_norm <- calcNormFactors(FilteredDGEList)
design <- model.matrix(~ VagNoPostcoptop32$phasecode + RIN + MalePresence)
colnames(design)[1:4] <- c("Intercept", "fertile", "postfertile", "prefertile")
voom_matrix <- voomWithQualityWeights(counts = DGE_counts_norm, design) # normalizes counts
dupcor <- duplicateCorrelation(voom_matrix, design, block=FemaleID)
fit <- lmFit(voom_matrix, design, block=FemaleID, correlation=dupcor$consensus)
fit<- eBayes(fit) # runs the linear model
```


#See results: fertile v nonfertile
```{r}
coeff<- as.data.frame(fit$coeff[,2])
pvalues<- as.data.frame(fit$p.value[,2])
qvalues<- as.data.frame(apply(fit$p.value, 2, function(x)p.adjust(x, method="fdr"))[,2])

FvNonF<-cbind(coeff, pvalues, qvalues)
colnames(FvNonF)<- c("coeff", "pvalue", "qvalue")
sig<-subset(FvNonF, qvalue < 0.1); dim(sig) 

Fertile_sig_genes<- sig
Fertile_sig_genes$Gene<-rownames(Fertile_sig_genes)
Fertile_sig_genes$comparison<-"FvNonF"
```

#prefertile v nonfertile
```{r}
coeff<- as.data.frame(fit$coeff[,4])
pvalues<- as.data.frame(fit$p.value[,4])
qvalues<- as.data.frame(apply(fit$p.value, 2, function(x)p.adjust(x, method="fdr"))[,4])

PreFvNonF<-cbind(coeff, pvalues, qvalues)
colnames(PreFvNonF)<- c("coeff", "pvalue", "qvalue")
sig<-subset(PreFvNonF, qvalue < 0.1); dim(sig) 

Prefertile_sig_genes<- sig
Prefertile_sig_genes$Gene<-rownames(Prefertile_sig_genes)
Prefertile_sig_genes$comparison<-"PreFvNonF"
```

#postfertile v nonfertile
```{r}
coeff<- as.data.frame(fit$coeff[,3])
pvalues<- as.data.frame(fit$p.value[,3])
qvalues<- as.data.frame(apply(fit$p.value, 2, function(x)p.adjust(x, method="fdr"))[,3])

PostFvNonF<-cbind(coeff, pvalues, qvalues)
colnames(PostFvNonF)<- c("coeff", "pvalue", "qvalue")
sig<-subset(PostFvNonF, qvalue < 0.1); dim(sig) 

Postfertile_sig_genes<- sig
Postfertile_sig_genes$Gene<-rownames(Postfertile_sig_genes)
Postfertile_sig_genes$comparison<-"PostFvNonF"
```


#make contrasts to get other pairwise comparisons
```{r}
contrast.matrix <- makeContrasts(
  prefertile_v_fertile = prefertile - fertile,
  prefertile_v_postfertile = prefertile - postfertile,
  fertile_v_postfertile= fertile - postfertile,
  levels = design
)

fit2 <- eBayes(contrasts.fit(fit, contrast.matrix))

```

#prefertile v fertile
```{r}

coeff<- as.data.frame(fit2$coeff[,1])
pvalues<- as.data.frame(fit2$p.value[,1])
qvalues<- as.data.frame(apply(fit2$p.value, 2, function(x)p.adjust(x, method="fdr"))[,1])

PreFvF<-cbind(coeff, pvalues, qvalues)
colnames(PreFvF)<- c("coeff", "pvalue", "qvalue")
PreFvF_sig<-subset(PreFvF, qvalue < 0.1); dim(PreFvF_sig) 
PreFvF_sig$Gene<-rownames(PreFvF_sig)
PreFvF_sig$comparison<-"PreFvF"
```

#prefertile v postfertile
```{r}
coeff<- as.data.frame(fit2$coeff[,2])
pvalues<- as.data.frame(fit2$p.value[,2])
qvalues<- as.data.frame(apply(fit2$p.value, 2, function(x)p.adjust(x, method="fdr"))[,2])

PreFvPostF<-cbind(coeff, pvalues, qvalues)
colnames(PreFvPostF)<- c("coeff", "pvalue", "qvalue")
PreFvPostF_sig<-subset(PreFvPostF, qvalue < 0.1); dim(PreFvPostF_sig) 
PreFvPostF_sig$Gene<-rownames(PreFvPostF_sig)
PreFvPostF_sig$comparison<-"PreFvPostF"
```

#fertile v postfertile
```{r}
coeff<- as.data.frame(fit2$coeff[,3])
pvalues<- as.data.frame(fit2$p.value[,3])
qvalues<- as.data.frame(apply(fit2$p.value, 2, function(x)p.adjust(x, method="fdr"))[,3])

FvPostF<-cbind(coeff, pvalues, qvalues)
colnames(FvPostF)<- c("coeff", "pvalue", "qvalue")
FvPostF_sig<-subset(FvPostF, qvalue < 0.1); dim(FvPostF_sig) 
FvPostF_sig$Gene<-rownames(FvPostF_sig)
FvPostF_sig$comparison<-"FvPostF"
```

#Write table of all results
```{r}
Allcomparisons<-do.call(rbind, list(Fertile_sig_genes, Prefertile_sig_genes, Postfertile_sig_genes, PreFvF_sig, PreFvPostF_sig, FvPostF_sig))
write.table(Allcomparisons, "DE_results_phase.txt", quote=F, row.names=F, col.names=T, sep="\t")
```

#visualize 
```{r}
## Heatmap of number of DE genes
DE_genes <- data.frame(V1=c("non-fertile", "non-fertile", "non-fertile", "pre-fertile", "pre-fertile", "fertile"), V2=c("pre-fertile", "fertile", "post-fertile", "fertile", "post-fertile", "post-fertile"), value=c(2390, 2481, 2111, 8, 58, 757))

ordered_phases<-c("non-fertile", "pre-fertile", "fertile", "post-fertile")

DE_genes$V1 <- factor(DE_genes$V1, levels = ordered_phases)
DE_genes$V2 <- factor(DE_genes$V2, levels = ordered_phases)

DEphase_heatmap<-ggplot(data=DE_genes, aes(x=V1, y=V2, fill=value)) +
  geom_tile(color="black") +
  geom_text(aes(label = value), size = 4) +
  scale_fill_gradient2(low = "white", high = "gray50") +
  theme_minimal() +
  coord_fixed() + 
  labs(x = "", y = "", fill = "DE genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=14, color="black"), axis.text.y = element_text(size=14, color="black"), legend.position = "right"); DEphase_heatmap

ggsave("DEphase_heatmap.pdf", DEphase_heatmap, width=4, height=4)
```

```{r}
#venn diagram

library(ggVennDiagram)

data <- list(
  Fertile = Fertile_sig_genes$Gene,
  "Pre-fertile" = Prefertile_sig_genes$Gene,
  "Post-fertile" = Postfertile_sig_genes$Gene
)

DE_venn<-ggVennDiagram(data, 
              label = "count", 
              set_color = c("#FEAE00","#FF644E", "#73BFFA"),
              set_size = 7,
              label_size = 6,
              edge_size = 2,
              category.names = c("Pre-fertile", "Fertile", "\nPost-fertile")) +
  scale_fill_gradient(low="white", high="white")+
    theme_void() +
  scale_x_continuous(expand = expansion(mult = .2)) +
  scale_y_continuous(expand = expansion(mult = .2)) +
  theme(legend.position="none"); DE_venn
    
ggsave("DE_venn.pdf", DE_venn, width=4, height=4)
```



################
# GSEA
################


```{r}
library(clusterProfiler)
library(enrichplot)
library(biomaRt)
```

```{r}
# Connect to Ensembl BioMart and select Papio anubis
ensembl <- useMart("ensembl", dataset = "panubis_gene_ensembl")  

# Retrieve GO term annotations for your genes
go_annotations <- getBM(attributes = c("external_gene_name", "go_id", "name_1006"),
                        mart = ensembl)

# Create TERM2GENE data frame (GO term to gene symbol mapping)
TERM2GENE <- go_annotations[, c("go_id", "external_gene_name")]

# Create TERM2NAME data frame (GO term to GO name mapping)
TERM2NAME <- unique(go_annotations[, c("go_id", "name_1006")])
```


```{r}
coeff_gsea<-as.data.frame(fit$coefficients)
coeff_gsea$geneID<-rownames(coeff_gsea)
```

#Fertile v non-fertile phase
```{r}
original_PhaseF_gene_list <- (coeff_gsea$fertile)
names(original_PhaseF_gene_list) <- coeff_gsea$geneID
PhaseF_gene_list<-na.omit(original_PhaseF_gene_list)
PhaseF_gene_list = sort( (PhaseF_gene_list), decreasing = TRUE)

```

```{r}
gsea_PhaseF <- GSEA(
  geneList = PhaseF_gene_list ,
  TERM2GENE = TERM2GENE,
  TERM2NAME = TERM2NAME,
  pvalueCutoff = 0.05
)
```

```{r}
write.table(gsea_PhaseF@result, "gsea_PhaseF.txt", quote=F, row.names=F, col.names=T, sep="\t")
```

#Visualize
```{r}
#dotplot
PhaseFdot<-dotplot(gsea_PhaseF, showCategory=20, color="p.adjust", split=".sign") + facet_grid(.~.sign) +
  theme(strip.text=element_text(size=15), axis.text.y=element_text(size=10, lineheight=0.7), axis.text.x=element_text(size=10)) +
  ggtitle("Fertile vs Non-fertile") +
  scale_fill_gradient(low= "gray30", high="gray100");PhaseFdot 
ggsave("FvNonFdot.pdf", PhaseFdot, height=8, width=8)
```

```{r}
#ridgeplot
PhaseFridge<-ridgeplot(gsea_PhaseF, showCategory=c("G protein-coupled receptor activity", "monoatomic ion channel complex", "positive regulation of inflammatory response", "phagocytic vesicle", "actin filament binding")) + labs(x = "enrichment distribution") + scale_fill_gradient(low= "gray30", high="gray100")+
  theme(axis.text.y=element_text(size=11, lineheight=0.8), axis.text.x=element_text(size=10)); PhaseFridge

ggsave("FvNonFdridge.pdf", PhaseFridge, height=3, width=6.5)
```


### Plot normalized counts of a couple genes!
```{r}
norm_counts<-voom_matrix$E
TLR2_counts<-as.data.frame(norm_counts["TLR2",])
colnames(TLR2_counts)<-c("norm_counts")
TLR2_counts$Sample_ID<-rownames(TLR2_counts)
TLR2_counts$gene<- "TLR2"
addphase<-VagNoPostcoptop32[,c("Sample_ID", "phase")]
TLR2_phase<-merge(TLR2_counts, addphase, by="Sample_ID")

TLR2_phase$phase<-factor(TLR2_phase$phase, levels=c("non-fertile", "pre-fertile", "fertile", "post-fertile"))

TLR2_plot<-ggplot(data=TLR2_phase, aes(x=phase, y=norm_counts)) +
  geom_boxplot(aes(color=phase), outlier.shape=NA) +
  geom_jitter(width=0.2, size=0.5, alpha=0.5)+
    scale_color_manual(values=c("#0076BA", "#FEAE00" , "#FF644E", "#73BFFA")) +
  theme_light() +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5, face="italic"), axis.text.x= element_text(angle=45, hjust=1, vjust=1, size=11)) +
  ylab("normalized counts") +
  xlab("") +
  ggtitle("TLR2"); TLR2_plot

ggsave("Phase_TLR2.pdf", TLR2_plot, width=2.3, height=3.5)
```


```{r}
norm_counts<-voom_matrix$E
SLC4A8_counts<-as.data.frame(norm_counts["SLC4A8",])
colnames(SLC4A8_counts)<-c("norm_counts")
SLC4A8_counts$Sample_ID<-rownames(SLC4A8_counts)
SLC4A8_counts$gene<- "SLC4A8"
addphase<-VagNoPostcoptop32[,c("Sample_ID", "phase")]
SLC4A8_phase<-merge(SLC4A8_counts, addphase, by="Sample_ID")

SLC4A8_phase$phase<-factor(SLC4A8_phase$phase, levels=c("non-fertile", "pre-fertile", "fertile", "post-fertile"))

SLC4A8_plot<-ggplot(data=SLC4A8_phase, aes(x=phase, y=norm_counts)) +
  geom_boxplot(aes(color=phase), outlier.shape=NA) +
  geom_jitter(width=0.2, size=0.5, alpha=0.5)+
    scale_color_manual(values=c("#0076BA", "#FEAE00" , "#FF644E", "#73BFFA")) +
  theme_light() +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5, face="italic"), axis.text.x= element_text(angle=45, hjust=1, vjust=1, size=11)) +
  ylab("normalized counts") +
  xlab("") +
  ggtitle("SLC4A8"); SLC4A8_plot

ggsave("Phase_SLC4A8.pdf", SLC4A8_plot, width=2.3, height=3.5)
```