---
title: "pH"
author: "Rachel"
date: "11/5/2021"
output: html_document
---

```{r}
library(lmerTest)
library(car)
library(sjPlot)
library(ggeffects)
library(ggplot2)
library(ggpubr)
library(tidyverse)
```

```{r}
pH_data<-read.csv("pH_data.csv")
```

#modeling effect of cycle phase on pH
```{r}
pH_noPostcop <- pH_data %>%
  filter(Postcop == "N")
pH_noPostcop$phase <- relevel(as.factor(pH_noPostcop$phase), ref = "non-fertile")

mod <- lmer(Avg.pH ~ phase + MalePresence + Avg.temp + (1|FemaleID), data=pH_noPostcop)
summary(mod)
```

```{r}
# Extract residuals
residuals <- resid(mod)

# Q-Q plot
qqnorm(residuals)

#colinearity
vif(mod)
```

#visualize
```{r}
dat <- ggpredict(mod, terms=c("phase"))
dat<- data.frame(dat)
dat.tidy <- dat %>%
  mutate(stder.high = (predicted + std.error)) %>%
  mutate(stder.low = (predicted - std.error))

pH_phase_plot <- ggplot(data= dat.tidy, aes(x=x, y=predicted)) +
  #geom_boxplot(aes(color=phase), outlier.shape=NA, size=0.5) +
  geom_point(aes(color=x),size=2) +
  geom_errorbar(aes(ymin=conf.low, ymax=conf.high, color=x), width=0.1, size=1) +
  scale_color_manual(values=c("#0076BA", "#FEAE00" , "#FF644E", "#73BFFA")) +
  geom_jitter(data= pH_phase_tidy, aes(x=phase, y=Avg.pH), width=0.25, size=0.3, alpha=0.2)+
  theme_light() +
  scale_x_discrete(labels = c("non-fertile", "pre-fertile", "fertile", "post-fertile")) +
  theme(plot.title= element_blank(), axis.text.y=element_text(size=12), axis.text.x=element_text(size=12, angle=45, hjust=1),  axis.title=element_text(size=14), legend.position="none") +
  xlab("") +
  ylab("Vaginal pH"); pH_phase_plot 

ggsave("pH_phase_plot.pdf", pH_phase_plot, width=3, height=3.5)
```


# post cop analyses

```{r}
Postcop_ph<-subset(pH_data, Postcop == "Y") #15
Noncop_ph<-subset(pH_data, Postcop == "N" & (phase == "pre-fertile" | phase == "fertile") & (FemaleID == "Suzon" | FemaleID == "Salade" | FemaleID == "Sottise" | FemaleID == "Vertige" | FemaleID == "Vuei")) #47
pH_postcop_tidy<-rbind(Postcop_ph, Noncop_ph)

mod2 <- lmer(Avg.pH ~ Postcop + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod2)
```

```{r}
# Extract residuals
residuals <- resid(mod2)

# Q-Q plot
qqnorm(residuals)

#colinearity
vif(mod2)
```

#visualize
```{r}
dat <- ggpredict(mod2, terms=c("Postcop"))
dat<- data.frame(dat)
dat.tidy <- dat %>%
  mutate(stder.high = (predicted + std.error)) %>%
  mutate(stder.low = (predicted - std.error))

pH_postcop_plot <- ggplot(data= dat.tidy, aes(y=predicted, x= x)) +
    geom_jitter(data= pH_postcop_tidy, aes(x=Postcop, y=Avg.pH), width=0.2, size=0.5, alpha=0.5)+
  geom_point(aes(color=x),size=2) +
  geom_errorbar(aes(ymin=conf.low, ymax=conf.high, color=x), width=0.1, size=1) +
  scale_color_manual(values= c("#531B93", "#D783FF")) +
  theme_light() +
  theme(plot.title= element_blank(), legend.position= "none") +
  scale_x_discrete(labels= c("N"= "No", "Y"= "Yes")) +
  xlab("Post-copulatory status") +
  ylab("Vaginal pH"); pH_postcop_plot 

ggsave("pH_postcop_plot.pdf", pH_postcop_plot, width=2.5, height=3)
```


#heteroscedasicity test
```{r}
library(lmtest)
mod2 <- lmer(Avg.pH ~ Postcop + phase + MaleStatus + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)

residuals <- resid(mod2)
fitted_values <- fitted(mod2)
lm_model <- lm(residuals ~ fitted_values)

#Breusch-Pagan test
bptest(lm_model)
```



#Female specific patterns figure 
```{r}
mean_data_grouped <- pH_postcop_tidy %>%
  group_by(FemaleID, Postcop) %>%
  summarize(MeanValue = mean(Avg.pH))

library(RColorBrewer)

pH_postcop_female_plot <- ggplot(data= pH_postcop_tidy, aes(x=Postcop, y=Avg.pH, color=FemaleID)) +
  geom_line(data= mean_data_grouped, aes(x=Postcop, y=MeanValue, group=FemaleID, color=FemaleID), size=1) +
  geom_jitter(data= pH_postcop_tidy, aes(x=Postcop, y=Avg.pH), width=0.1, size=0.5)+
  theme_light() +
  scale_color_brewer(palette = "Set1", labels=c(1:5), name="Female ID")+
  scale_x_discrete(labels = c("No", "Yes")) +
  theme(plot.title= element_blank()) +
  xlab("Post-copulatory Status") +
  ylab("Vaginal pH"); pH_postcop_female_plot 

ggsave("pH_postcop_perFemale.pdf", pH_postcop_female_plot, width=3.5, height=3)
```


### post cop genotype analyses ###

#stMLH
```{r}
mod7 <- lmer(Avg.pH ~ Postcop*Male_stMLH + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)

summary(mod7)
mod7Anova <- Anova(mod7); mod7Anova 
mod7pvalue <- mod7Anova[6,3]

plot_model(mod7, type="pred", terms=c("Postcop", "Male_stMLH"))
```


# Male MHC I diversity
```{r}
mod8 <- lmer(Avg.pH ~  Postcop*Male_MHCI_allele + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)

summary(mod8)
mod8Anova <- Anova(mod8); mod8Anova
mod8pvalue <- mod8Anova[6,3]
plot_model(mod8, type="pred", terms=c("Postcop", "Male_MHCI_allele"))
```

#Male MHCII allele diversity
```{r}
mod9 <- lmer(Avg.pH ~  Postcop*Male_MHCII_allele + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod9)
mod9Anova <- Anova(mod9); mod9Anova
mod9pvalue <- mod9Anova[6,3]
plot_model(mod9, type="pred", terms=c("Postcop [all]", "Male_MHCII_allele [all]"))
```


#Male MHCI supertype diversity
```{r}
mod10 <- lmer(Avg.pH ~Postcop*Male_MHCI_supertype + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod10)
mod10Anova <- Anova(mod10); mod10Anova
mod10pvalue <- mod10Anova[6,3]
plot_model(mod10, type="pred", terms=c("Postcop [all]", "Male_MHCI_supertype [all]"))
```

#Male MHCII supertype diversity
```{r}
mod11 <- lmer(Avg.pH ~  Postcop*Male_MHCII_supertype + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod11)
mod11Anova <- Anova(mod11); mod11Anova
mod11pvalue <- mod11Anova[6,3]
plot_model(mod11, type="pred", terms=c("Postcop", "Male_MHCII_supertype"))
```

```{r}
p.adjust(c(mod7pvalue, mod8pvalue, mod9pvalue, mod10pvalue, mod11pvalue), method="BH")
```



## complementarity analyses

#KINSHIP
```{r}
mod12 <- lmer(Avg.pH ~  Postcop*KINSHIP + phase + MalePresence + Avg.temp + (1|MaleID)+ (1|FemaleID), data=pH_postcop_tidy)

summary(mod12)
mod12Anova <- Anova(mod12); mod12Anova
mod12pvalue <- mod12Anova[6,3]
```

```{r}
dat <- ggpredict(mod12, terms=c("KINSHIP [all]", "Postcop"))
dat<- data.frame(dat)
dat.tidy <- dat %>%
  mutate(stder.high = (predicted + std.error)) %>%
  mutate(stder.low = (predicted - std.error))

KIN_pH_oneplot <- ggplot(data= dat.tidy, aes(y=predicted, x= group, color=x, group=x)) +
  geom_point(position=position_dodge(width=0.8), size=3) +
  geom_errorbar(aes(ymin = stder.low, ymax = stder.high, linetype=NULL, group=x), width=0.3, padding=0.3, position=position_dodge(width=0.8), size=1) +
    theme_bw() +
  theme(plot.title=element_text(size=16, hjust=0.5), axis.title=element_text(size=12)) +
    geom_point(data= na.omit(pH_postcop_tidy), aes(x=Postcop, y=Avg.pH, color=KINSHIP, group=KINSHIP), size=2, shape= 21, position=position_dodge(width=0.8)) + 
   scale_color_gradient2(low = "grey1", mid= "grey40", high = "grey80", name= "Kinship", trans= "reverse", limits= c(0.25, -0.25), breaks= c(0.2, 0.1, 0, -0.1, -0.2), labels= c(0.2, 0.1, 0, -0.1, -0.2))+
  ylab("Vaginal pH") +
  xlab("Post-copulatory Status") +
  ggtitle("Kinship") +
  scale_x_discrete(labels=c("No", "Yes"));KIN_pH_oneplot

ggsave("Kinship_pH.pdf", KIN_pH_oneplot, width=4, height=4)
```


#MHC I allele comp

```{r}
mod13 <- lmer(Avg.pH ~  Postcop*MHCI_allele_comp + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod13)
mod13Anova <- Anova(mod13); mod13Anova
mod13pvalue <- mod13Anova[6,3]
plot_model(mod13, type="pred", terms=c("Postcop", "MHCI_allele_comp"))
```

MHCII allele comp
```{r}
mod14 <- lmer(Avg.pH ~  Postcop*MHCII_allele_comp + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod14)
mod14Anova <- Anova(mod14); mod14Anova
mod14pvalue <- mod14Anova[6,3]
plot_model(mod14, type="pred", terms=c("Postcop", "MHCII_allele_comp"))
```


```{r}
dat <- ggpredict(mod14, terms=c("MHCII_allele_comp [all]", "Postcop"))
dat<- data.frame(dat)
dat.tidy <- dat %>%
  mutate(stder.high = (predicted + std.error)) %>%
  mutate(stder.low = (predicted - std.error))


MHCII_allele_comp_pH_oneplot <- ggplot(data= dat.tidy, aes(y=predicted, x= group, color=x, group=x)) +
  geom_point(position=position_dodge(width=0.6), size=3) +
  geom_errorbar(aes(ymin = stder.low, ymax = stder.high, linetype=NULL, group=x), width=0.3, position=position_dodge(width=0.6), size=1) +
    theme_bw() +
  theme(plot.title=element_text(size=16, hjust=0.3), axis.title=element_text(size=12)) +
    geom_point(data= na.omit(pH_postcop_tidy), aes(x=Postcop, y=Avg.pH, color=MHCII_allele_comp, group=MHCII_allele_comp), size=1.8, shape=21, position=position_dodge(width=0.6)) +
   scale_color_gradient2(high = "grey1", mid= "grey40", low="grey80", midpoint= 0.5, name= "Proportion \nshared \nalleles", breaks=c(0,0.2,0.4,0.6,0.8,1.0), limits=c(-0.05, 1.05)) +
  guides(color = guide_colorbar(reverse=T)) +
  ylab("Vaginal pH") +
  xlab("Post-copulatory Status") +
  ggtitle("MHC II allelic complementarity") +
  scale_x_discrete(labels=c("No", "Yes"));MHCII_allele_comp_pH_oneplot

ggsave("MHCII_allele_comp_pH.pdf", MHCII_allele_comp_pH_oneplot, width=4, height=4)
```



#MHCI supertype comp
```{r}
mod15 <- lmer(Avg.pH ~  Postcop*MHCI_supertype_comp + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod15)
mod15Anova <- Anova(mod15); mod15Anova
mod15pvalue <- mod15Anova[6,3]
plot_model(mod15, type="pred", terms=c("Postcop", "MHCI_supertype_comp"))
```


#MHCII supertype comp
```{r}
mod16 <- lmer(Avg.pH ~  Postcop*MHCII_supertype_comp + phase + MalePresence + Avg.temp + (1|MaleID) + (1|FemaleID), data=pH_postcop_tidy)
summary(mod16)
mod16Anova <- Anova(mod16); mod16Anova
mod16pvalue <- mod16Anova[6,3]
plot_model(mod16, type="pred", terms=c("Postcop", "MHCII_supertype_comp"))
```

```{r}
dat <- ggpredict(mod16, terms=c("MHCII_supertype_comp [all]", "Postcop"))
dat<- data.frame(dat)
dat.tidy <- dat %>%
  mutate(stder.high = (predicted + std.error)) %>%
  mutate(stder.low = (predicted - std.error))


MHCII_supertype_comp_pH_oneplot <- ggplot(data= dat.tidy, aes(y=predicted, x= group, color=x, group=x)) +
geom_point(position=position_dodge(width=0.8), size=3) +
  geom_errorbar(aes(ymin = stder.low, ymax = stder.high, linetype=NULL, group=x), width=0.3, position=position_dodge(width=0.8), size=0.7) +
    theme_bw() +
  theme(plot.title=element_text(size=16, hjust=0.3), axis.title=element_text(size=12)) +
    geom_point(data= na.omit(pH_postcop_tidy), aes(x=Postcop, y=Avg.pH, color=MHCII_supertype_comp, group=MHCII_supertype_comp), size=1.5, shape=21, position=position_dodge(width=0.8)) +
   scale_color_gradient2(high = "grey0", mid= "grey40", low="grey80", midpoint= 0.5, name= "Proportion \nshared \nsupertypes", breaks=c(0,0.2,0.4,0.6,0.8,1.0), limits=c(-0.05, 1.05)) +
  guides(color = guide_colorbar(reverse=T)) +
  ylab("Vaginal pH") +
  xlab("Post-copulatory Status") +
  ggtitle("MHC II supertype complementarity") +
  scale_x_discrete(labels=c("No", "Yes"), expand=expansion(mult=.2));MHCII_supertype_comp_pH_oneplot

ggsave("MHCII_supertype_comp_pH.pdf", MHCII_supertype_comp_pH_oneplot, width=4, height=4)
```


```{r}
p.adjust(c(mod12pvalue, mod13pvalue, mod14pvalue, mod15pvalue, mod16pvalue), method="BH")
```


## Plot 3 significant pH 

```{r}
Three_pH_plots<- ggarrange(KIN_pH_oneplot, MHCII_allele_comp_pH_oneplot, MHCII_supertype_comp_pH_oneplot, labels=c("a)", "b)", "c)"), hjust=-0.2, nrow=1, ncol=3); Three_pH_plots
```


